---
title: "STAT545 HW01"
author: "Ernest Lai"
date: "September 17, 2018"
output: html_document
---

<br/>

# R Markdown

This is the R Markdown document for hw01, which explores the **gapminder** dataset. To explore the data, we first have to install the 'gapminder' package through **install.packages('gapminder')** and open the dataset into our R-Studio environment through **library(gapminder)**.

<br/>
*For additional information on functions used in this exercise, type "?" and the function as one word in the console and the R help documentation will explain the function and the object needed to run the function successfully*
 

## Viewing and Exploring the dataset

To first explore contents of the dataset, we should assign our gapminder dataset to a variable called 'data'. Prior to this, please be reminded to do a one-time installation of gapminder dataset via **install.packages('gapminder')**:

<br/>

```{r view, message=FALSE, warning=FALSE}

library(gapminder)
data <- gapminder

```
*The statements message=FALSE and warning=FALSE excludes external warning messages or output messages from being printed in the HTML version*

<br/>

From this point on, any work associated with gapminder in this R-Studio session will be referenced to the variable called 'data'. Lets try exploring the gapminder dataset:

<br/>
```{r explore, message=FALSE, warning=FALSE, tidy=FALSE}

 # Provides number of observations and variables, also the class of each variable in the gapminder dataset
str(data)

# Information provided for each variable for the first 6 observed entries in the gapminder dataset
head(data) 

# Number of rows in gapminder dataset (or number of observations)
nrow(data) 

# Number of columns in gapminder dataset (or number of variables)
ncol(data) 

```

<br/>

## Summary Statistics
We can produce summary statistics for specific variables within the gapminder dataset, by adding the syntax '$' after the gapminder dataset name (data), followed by the variable name in the gapminder dataset

<br/>
```{r summary, message=FALSE, warning=FALSE}

# Gives the frequency of each country in the dataset
table(data$country) 

# Gives the frequency of each continent in the dataset
table(data$continent) 

# Number of countries in the data set
length(
  table(data$country)
  ) 

# Number of continents in the data set
length(
  table(data$continent)
  ) 

# Gives the average population based on all the countries in the data set
mean(data$pop) 

# Gives the variance of the population based on all countries in the data set
var(data$pop) 

# Gives the 1st quartile, median, third quartile, minimum, maximum, and average values for GDP per capita 
summary(data$gdpPercap) 

```

<br/>

## Summary Statistics with Specific Requirements

As seen in the dataset, there are a lot different countries being observed.  What if we are interested in the average or range of the population for a specific country? Or GDP per capita given a certain continent? Or even average population with a given range of GDP per capita?

<br/> 

```{r specifics, message = FALSE, warning=FALSE}

# Takes the average population for Montenegro (average population, given country = Montenegro)
mean(
  data$pop[data$country == "Montenegro"]
  ) 

# Max, min, mean, median, Q1, and A3 values for population in the Asia continent
summary(
  data$pop[data$continent == "Asia"]
  ) 

# Max, min, mean, median, Q1, and A3 values for GDP per capita in the Europe continent
summary(
  data$gdpPercap[data$continent == "Europe"]
  ) 

# Mean population of countries with GDP per capita between 1202.1 and 3531.8 inclusive.
mean(
  data$pop[data$gdpPercap >= '1202.1' & data$gdpPercap <= '3531.8']
  ) 
```

<br/>

## Including Plots

Plots can also be made for each variable. Which plot we use is dependent on the type of data we are dealing with (ie. continuous, discrete, etc)

<br/>
```{r  plots, message = FALSE, warning = FALSE}
# Boxplot of population for Morocco
boxplot(
  data$pop[data$country == 'Morocco'], 
  ylab = 'Population', xlab ='Country', main = 'Frequency of Each Continent in gapminder Dataset'
  )

# Boxplot of life expectancy for each continent
boxplot(data$lifeExp ~ data$continent, 
        xlab = 'Continent', ylab = 'Life Expectancy (by Years)', main = 'Boxplot of Life Expectancy (in
        Years) by Continent'
        ) 

# Histogram of Life Expectancy for the African Continent
hist(data$lifeExp[data$continent =='Africa'], 
     xlab='Life Expectancy (in Years)', ylab='Frequency', main = 'Life Expectancy (in Years) for Africa)'
     )
```

